---
title: 'W203 Statistics for Data Science: Lab 1'
author: "James Beck & Ted Pham"
date: "January, 2017"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; With the data provided, an exploratory data analysis was performed in an attempt to uncover a connection between CEO salary and company performance. Additional observations and relationships observed between other variables of interest were also explored. This report explains and visualizes the process by which these connections were uncovered and provides commentary around the observed connections in an attempt to contextualize them within the dataset. 

The variables that we have the ability to consider in our analysis include the following:
```{r load}
setwd("/Users/Beck/UCBMIDS/stats/lab_1")
load("ceo_w203.RData")
names(CEO)
```

The questions we can ask from the 8 variables are:  

1. Does CEO salary correlate with profits? How does this change when normalizing profits/market value? (plotted will be added)
2. Does having a degree and even a graduate degree help the CEO perform the job?
3. How does CEO tenure at the company and as CEO affect profits?
4. Do younger CEOs have more or less comten & ceoten and does age correlate with profits?
5. Are the CEOs being paid accordingly to the company market value?
  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; From answering these questions, we can form a statistical model on how to adequately compensate for a CEO's company tenure, market value, profits, education, etc. so that we can properly motivate the CEO in the job. However, to preface the analysis, it is important to remember that CEOs typically earn compensation through means other than simply their salary such as stock option and bonuses that are not presented in this dataset. Therefore, to make a more accurate model, these factors would need to be considered. 

## Data Check

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; We first would like to check out the dataset from a high level to look for any missing values, NAs, or other data characteristics that could create issues during later analysis. We first check to see if the dataset contains any NAs that will need to be handled appropriately to avoid any complications with calculation. We will also check to make sure that all variables have the same number of values to ensure that we are working with variables of equal size (i.e. no missing values):  
&nbsp;
```{r}
sapply(CEO, function(x) sum(is.na(x)))
sapply(CEO, function(x) length(x))
```
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Since the output of this NA check is zero for all variables in the dataset we can conclude that NAs do not need to be a further consideration while we continue to explore and analyze the data. Additionally, all of the variable lengths appear to be the same size so we do not appear to be missing any data.

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; As another first step, we will check the summary of each of the variables to get an initial sense of the shape of the data we will be analyzing and attempt to note anything of immediate interest: 
&nbsp;
```{r}
summary(CEO)
```
&nbsp; 
```{r}
library(car)
scatterplotMatrix(CEO, diagonal = "histogram")
```
&nbsp;  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; From the summary we can make some cursory conclusions about the nature of the "average CEO" in 1990. We notice that the CEO age follows a nice normal distribution with most CEOs falling in 50-60 years old age bracket. All of the CEOs in the data set worked for the company for at least two years before being promoted to CEO. In other words, no CEO in our data set was hired from outside of the company. 21 years with the company and 7 years as CEO comprise the median experience. As mentioned in the intrduction, means of compensation other than salary are missing from the data set so all analysis will be conducted based on salary as the sole source of compensation for the CEOs.
&nbsp;  
&nbsp; 
**A few notable observations from this summary and the scatterplot matrix:**


1. **Salary** -  A 3rd quarratile of 1101 with a max value of 5299 seems to imply the existence of at least one notable outlier in the set. A histogram of the values confirms and highlights the existence of this outlier as well as showing the overall right skewness of the data.  
&nbsp;
```{r}
hist(CEO$salary, breaks = seq(0,6000,by=100),
     main = "Histogram of CEO Salary",
     xlab = "CEO Salary $Th")
```

2. **College** - This variable appears to be binary (college or no college) where `r round(mean(CEO$college)*100,digits=2) `% of CEOs went to college. Additionally upon closer inspection it appears that there are only a very small number of CEOs in the set that did not attend college so making any meaningful conclusions about the effects that college attendance has may not be worthwhile.

```{r}
library(ggplot2)
ggplot(CEO, aes(college)) + geom_histogram(bins=3)
```

3. **Grad** -  The variable also appears to be binary (grad school or no grad school) where `r round(mean(CEO$grad)*100,digits=2) `% of CEOs went to grad school. Unlike the college variable, there seems to be a healthy split in the data set between the number of CEOs that attended graduate school and those that didn't.
```{r}
library(ggplot2)
ggplot(CEO, aes(grad)) + geom_histogram(bins=3)
```
```{r}
subset(CEO, college == 0 & grad ==1)
```
Interestingly, as seen above, there are two entries where the CEOs did not attend college but did attend graduate school. Although this might seem odd, some MBA schools do accept students without a college degree, hence we will not disclude these in our subsequent analysis.

&nbsp;  
4. **Profits** - Noting that the minimum of the set is `r min(CEO$profits)` we will have to keep in mind that not all companies are posting a positive profit for the business. Additionally, the maximum of value of `r max(CEO$profits)` seems notably higher than the 3rd quartile value as visualized in the histogram below.  
&nbsp;
```{r}
hist(CEO$profits, breaks = seq(-500,3000,by=50),
     main = "Histogram of Company Profits",
     xlab = "Profits $M")
```
&nbsp;  
&nbsp;  
5. **Market Value** -  Here we have a problem. The minimum market value of -1 is a red flag that upon further investigation shows what is apparently some problematic entries. We can confine the data set to just those entires with a market value of -1 to investigate further:
&nbsp;  
```{r}
CEO_neg_val = subset(CEO, mktval == -1)
sapply(CEO_neg_val, function(x) length(x))
head(CEO_neg_val)
```
  
What we can see here is that there are 5 entries that have this issue and in addition to having a market value of -1 these entries also have a profit value of exactly -1. Based on this observation these entries appear to be mistakes and will not be included in further analysis.  
&nbsp;
```{r}
CEO_fixed = subset(CEO, mktval != -1)
```

## CEO Salary & Company Profits

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; To start, we can first explore the primary relationship of interest in the data set: CEO salary and company profits.

```{r}
library(car)
r_string = paste("r =",round(cor(CEO_fixed$profits, CEO_fixed$salary),3))
plot(CEO_fixed$salary, CEO_fixed$profits,
     main = "Company Profits vs. CEO Salary",
     xlab = "CEO Salary $Th",
     ylab = "Company Profits $M")
abline(lm(CEO_fixed$profits ~ CEO_fixed$salary))
text(5000,1000,r_string)
```
&nbsp;  
&nbsp;  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; As seen in the scatterplot above with the corresponding regression line and printed r value, there is a positive correlation between CEO salary and company profits. However, the r value is only 0.399 and it's easy to see why from the plot. There are several values that are quite high on the CEO pay scale with very low or even negative profits for the company.
&nbsp;  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; We also ask if there is a difference in the mean salary of CEOs from companies that received positive profit vs negative profit in 1990. For this analysis we conducted a student t-test between the the means of the CEO salaries of both subsets of the sample. Our p-value of 0.9774 from the student t-test indicates that there was no difference in mean salary between negative and positive profit companies. The boxplot corroborates the conjecture. The distribution of salaries for CEOs that achieved negative profits fo their companies is notably tighter than the distribution for positive profit CEOs, but this could likely be a result of the much smaller subset size for negative profit CEOs. This leads us to further hypothesize that CEO's base salary is not exclusively determined with the current year profit nor does salary the only one determining factor that drives profit.

```{r}
CEOpos = subset(CEO_fixed, profits > 0)
CEOpos$profit = 1
CEOneg = subset(CEO_fixed, profits <= 0)
CEOneg$profit = 0
CEOprofit = merge(CEOpos, CEOneg, all = TRUE)
boxplot(salary ~ profit, data = CEOprofit, names = c("Negative Profit","Positive Profit"),
        ylab = "CEO Salary in $Th")
t.test(CEOpos$salary,CEOneg$salary)
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; We could stop here and conclude with the simple observation that there does in fact appear to be some mild relationship between CEO salary and company profit, but we would be missing out on exploring several other potentially interesting variables and relationships within this data set.

## Salary and Market Value
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Regardless of the company profit which might have negative values, we suspect CEO salary is related to the company market value, an indicator of the company size. To test this, we plot salary vs market value. From the plots, we can discern that CEO salary increases linearly with the logarithmic market value. There seems to be a ceiling for CEO pay from the sampled data. In addition, the highest paid CEO is from a \$1.5B company whereas the CEO of the highest market value company (\$45B) only earned $2M. We again hypothesize that this could be due to the data set's lack of compensation information beyond salary, but without that information available can make no firm conclusion.

```{r}
par(mfrow=c(1,2))
CEOmkt = CEO_fixed
plot(CEOmkt$mktval,CEOmkt$salary,
     main = "CEO Salary vs. Market Value",
     ylab = "CEO Salary $Th",
     xlab = "Market Value $M")

plot(CEOmkt$mktval,CEOmkt$salary, log = "x",
     main = "CEO Salary vs. Market Value",
     ylab = "CEO Salary $Th",
     xlab = "Market Value $M (log scale)")
```
```{r}
CEOmkt$logvalue = CEOmkt$mktval
cor(CEOmkt$salary,CEOmkt$logvalue)
```
The relationship between salary and log(market value) is moderately positive with a correlation of 0.4082068.


## College and Grad School

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Since we are still interested in how variables in the data set affect company profits, let's examine that metric from a different angle: college and graduate school attendance.
&nbsp;  
```{r}
boxplot(profits ~ college, data=CEO_fixed, names=c("No College","Collge"),
        main = "Company Profits vs. CEO College Attendance",
        ylab="Company Profits $M")
```
&nbsp;  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; It is interesting to see that the population of CEOs that did not attend college did not produce any of the negative profit companies. However, looking to the size of the 'No College' subset we see the following:
```{r}
length(subset(CEO_fixed, college == 0))
```

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Unfortunately, there are only 8 entries in the data set that represent CEOs that have no college education. Since this is such a small set we do not feel we can make any meaningful conclusions based on this subset of the population. Comparing CEOs that have attended graduate school and those that have proves more interesting though as both of those subsets have a healthy number of entries to pull from.

```{r}
boxplot(profits ~ grad, data=CEO_fixed, names=c("No Grad School","Grad School"),
        main = "Company Profits vs. CEO Grad School Attendance",
        ylab="Company Profits $M")
```
&nbsp;  
&nbsp;  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; There are few small points to be taken from this box plot - The grad school quartiles are somewhat more spread out that the 'no grad' quartiles. Each set contains some outliers on the high end and very few low end outliers. If we look to scatter plots and regression lines for each subset, however, we can see something interesting.
&nbsp;  
&nbsp;  
```{r}
CEO_grad = subset(CEO_fixed, grad == 1)
r_string = paste("r =",round(cor(CEO_grad$profits, CEO_grad$salary),3))
plot(CEO_grad$salary, CEO_grad$profits,
     main = "Company Profits vs. CEO Salary w/ graduate school",
     xlab = "CEO Salary $Th",
     ylab = "Company Profits $M")
abline(lm(CEO_grad$profits ~ CEO_grad$salary))
text(2000,1000,r_string)

```

```{r}
CEO_no_grad = subset(CEO_fixed, grad == 0)
r_string = paste("r =",round(cor(CEO_no_grad$profits, CEO_no_grad$salary),3))
plot(CEO_no_grad$salary, CEO_no_grad$profits,
     main = "Company Profits vs. CEO Salary w/o graduate school",
     xlab = "CEO Salary $Th",
     ylab = "Company Profits $M")
abline(lm(CEO_no_grad$profits ~ CEO_no_grad$salary))
text(4000,1000,r_string)

```
&nbsp;  
&nbsp;  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Apparently, as evidenced by the differing correlation coefficients, salary is a much better indicator of company profit for companies with a CEO who has graduate school experience than for companies where the CEO does not have graduate school experience.

## Do Age and Experience Correlate with Salary?

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Our boxplots do not reveal any obvious relationship between age or level of experience and salary. We did however found another error in the data where one CEO has more CEO tenure than company tenure. 
```{r}
par(mfrow= c(3,1))
boxplot(salary ~ age, data = CEO_fixed,
        main = " Salary & Age",
        xlab = "Age (Years)",
        ylab = "Salary in $Th")

boxplot(salary ~ comten, data = CEO_fixed,
        main = " Salary & Yr with Comp",
        xlab = "Company Tenure (Years)",
        ylab = "Salary in $Th")

boxplot(salary ~ ceoten, data = CEO_fixed,
        main = " Salary & Yr as CEO",
        xlab = "CEO Tenure (Years)",
        ylab = "Salary in $Th")



```


## Discussion & Conclusion

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; 

Errors in the data. No information whether the data set was randomized or exclusive so cannot make inference about 1990 population. 

Key Findings from the exploratory analysis:
1.	Profit has a moderate positive linear relationship with salary.
2.	Salary is linearly correlated with the lograthmic market value.
3.	Grad school vs non grade
4.	Age and level of experience do not correlate with salary


Confounding effects:
Other modes of compensation for the CEO. Business sector can also influence the profit because two similarly valued company but belong to two different business sectors might perform differently and thus have different profits.  Business sector can also affect the requirement of CEO education.


??? other analyses to include if there are still space
???Ceoten  = comten -> startup? Founders more stakes?
???How did the economy grow in 1990? Average profit/mktval histogram

